---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# slcflights

<!-- badges: start -->
<!-- badges: end -->

The goal of slcflights is to provide packaged Parquet files for Salt Lake 
City-related U.S. flight records, along with a CSV table of airport coordinates 
used to support those records.

The packaged flight data are derived from the U.S. commercial flight data used
in the 2025 ASA Data Expo Challenge. The underlying flight records originate
from the Bureau of Transportation Statistics (BTS) TranStats On-Time
Performance data, and the packaged coordinate table is derived from the BTS
TranStats Master Coordinate support table.

The package includes two flight-data groupings:

* **main records**: flights where Salt Lake City appears in the core origin or
  destination airport fields
* **diversion-only records**: flights where Salt Lake City appears only in
  diversion airport fields and not in the core origin or destination airport
  fields

## Installation

You can install the development version of slcflights like so:

``` r
# install.packages("remotes")
remotes::install_github("AlKulAK/slcflights")
```

## Package contents

The main user-facing functions are

* `available_years()` to list packaged years for main or diversion-only records
* `read_year_main()` to read one year's main flight records into memory
* `read_year_div()` to read one year's diversion-only flight records into memory
* `read_main()` to read one, many, or all years of main records into memory
* `read_div()` to read one, many, or all years of diversion-only records into
  memory
* `open_main()` to open one, many, or all years of main records lazily as an
  Arrow dataset
* `open_div()` to open one, many, or all years of diversion-only records lazily
  as an Arrow dataset
* `read_coords()` to read the packaged airport coordinate table

## Examples

Load the package:

```{r}
library(slcflights)
```

List the packaged years available for main records:

```{r}
available_years("main")
```

Read the airport coordinate table:

```{r}
coords <- read_coords()
head(coords)
```

Read one year's main flight records into memory:

```{r}
x_1987 <- read_year_main(1987)
head(x_1987)
```

Read multiple years of main records into memory:

```{r}
x_main <- read_main(1987:1988)
dim(x_main)
```

Open main records lazily with Arrow when you want to defer reading until later:

```{r}
ds_main <- open_main(1987:1988)
ds_main
```

Read diversion-only records for one year:

```{r, eval = FALSE}
x_div <- read_year_div(2015)
head(x_div)
```

## Rebuilding the packaged data

The packaged Parquet files and coordinate CSV can be rebuilt from source data.

The rebuild workflow is implemented in `data-raw/build_data.R`. 
To rebuild, run from a source checkout of the package:

```r
source("data-raw/build_data.R")
build_slc_data()
```

By default, `build_slc_data()`

* rebuilds data for years 1987 through 2024
* uses Salt Lake City's BTS airport ID (`14869`)
* downloads source Parquet files into `data-raw/cache/`
* writes the reduced coordinate CSV into the build cache
* copies final packaged outputs into `inst/extdata/`
* deletes the build cache when finished

Note that you can rebuild a subset of years, for example:

```r
source("data-raw/build_data.R")
build_slc_data(years = 2019:2024)
```

Because rebuilding requires large downloads and substantial disk space, it is recommended to work with the packaged files directly.

## Data provenance

This package repackages a Salt Lake City-focused subset of the flight data used
in the 2025 ASA Data Expo Challenge. The challenge centered on analysis and
visualization of U.S. commercial flight arrival and departure records, while
the underlying data ultimately come from BTS TranStats On-Time Performance
data.

The packaged coordinate table is derived from the BTS TranStats Master
Coordinate support table and contains the airport sequence identifiers needed
to enrich the flight records with airport latitude, longitude, and date-bounded
airport metadata.
