---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# slcflights

<!-- badges: start -->
<!-- badges: end -->

The goal of slcflights is to provide packaged Parquet files for Salt Lake
City-related U.S. flight records, along with a CSV table of airport coordinates
used to support those records.

The packaged flight data are derived from the U.S. commercial flight data used
in the 2025 ASA Data Expo Challenge. The underlying flight records originate
from the Bureau of Transportation Statistics (BTS) TranStats On-Time
Performance data, and the packaged coordinate table is derived from the BTS
TranStats Master Coordinate support table.

The package includes two flight-data groupings:

* **main records**: flights where Salt Lake City appears in the core origin or
  destination airport fields
* **diversion-only records**: flights where Salt Lake City appears only in
  diversion airport fields and not in the core origin or destination airport
  fields

The installed package data cover **October 1, 1987 through June 30, 2024**.
This means that 1987 and 2024 are partial years. For analyses requiring
complete calendar years using only installed data, the cleanest full-year span
is 1988 through 2023.

Users can optionally extend the data beyond June 2024 by building a local
cache with `update_slcflights_data()`. The installed package files are never
modified.

## Installation

You can install the development version of slcflights like so:

``` r
# install.packages("remotes")
remotes::install_github("AlKulAK/slcflights")
```

## Package contents

The main user-facing functions are

* `available_years()` to list available years for main or diversion-only
  records
* `read_year_main()` to read one year's main flight records into memory
* `read_year_div()` to read one year's diversion-only flight records into
  memory
* `read_main()` to read one, many, or all years of main records into memory
* `read_div()` to read one, many, or all years of diversion-only records into
  memory
* `open_main()` to open one, many, or all years of main records lazily as an
  Arrow dataset
* `open_div()` to open one, many, or all years of diversion-only records lazily
  as an Arrow dataset
* `read_coords()` to read the airport coordinate table
* `update_slcflights_data()` to download newer BTS data into a local user cache
* `slcflights_cache_info()` to inspect the local user cache
* `clear_slcflights_cache()` to remove the local user cache

The reader functions are cache-aware. If no local cache is active, they read the
installed package data. If a local cache is active, they read the installed data
together with compatible cached data.

## Examples

Load the package:

```{r}
library(slcflights)
```

List the available years for main records:

```{r}
available_years("main")
```

Read the airport coordinate table:

```{r}
coords <- read_coords()
head(coords)
```

Read one year's main flight records into memory:

```{r}
x_1987 <- read_year_main(1987)
head(x_1987)
```

Read multiple years of main records into memory:

```{r}
x_main <- read_main(1987:1988)
dim(x_main)
```

Open main records lazily with Arrow when you want to defer reading until later:

```{r}
ds_main <- open_main(1987:1988)
ds_main
```

Read diversion-only records for one year:

```{r, eval = FALSE}
x_div <- read_year_div(2015)
head(x_div)
```

## Updating local data

The installed package includes flight records through June 2024. To extend the
available data beyond the installed endpoint, use:

```{r, eval = FALSE}
update_slcflights_data()
```

By default, `update_slcflights_data()` checks BTS for the latest available
monthly data and downloads every available month after June 2024. You can also
request an explicit endpoint:

```{r, eval = FALSE}
update_slcflights_data(until = "2024-07")
update_slcflights_data(until = c(2024, 12))
```

Local updates always begin with July 2024 and form a consecutive extension of
the installed data.

The local update workflow

* downloads monthly BTS On-Time Performance files into the user cache
* downloads the BTS Master Coordinate table into the user cache
* builds annual main and diversion-only Parquet files in a staging directory
* reduces the coordinate table to airports used by installed and cached records
* enriches cached Parquet files with coordinate metadata
* finalizes cached Parquet schemas so they match the installed schema contract
* activates the completed cache atomically

The installed package files are not changed. The local cache is stored in the
user cache directory returned by `tools::R_user_dir("slcflights", "cache")`.

Inspect the active cache with:

```{r, eval = FALSE}
slcflights_cache_info()
```

After a cache is active, the ordinary readers automatically use it. For example,
after updating through July 2024, `read_year_main(2024)` reads both the
installed January-through-June records and the cached July records.

Remove the local cache with:

```{r, eval = FALSE}
clear_slcflights_cache()
```

## Rebuilding the packaged data

The packaged Parquet files and coordinate CSV can be rebuilt from source data.

This is a maintainer workflow implemented in `data-raw/build_data.R`. To
rebuild, run from a source checkout of the package:

```r
source("data-raw/build_data.R")
build_slc_data()
```

By default, `build_slc_data()`

* rebuilds data for years 1987 through 2024
* uses Salt Lake City's BTS airport ID (`14869`)
* downloads the annual source Parquet files into a temporary build directory
* removes columns that are globally all `NULL`
* filters the data to Salt Lake City-related records
* removes columns that are all `NULL` after Salt Lake City filtering
* sorts rows by flight date and scheduled departure time
* splits the filtered data into main and diversion-only files
* reduces the BTS Master Coordinate table to the airport sequence IDs used by
  the packaged flight data
* enriches the final Parquet files with airport coordinate metadata
* copies final packaged outputs into `inst/extdata/`
* removes the temporary build directory when finished

The temporary build directory is a maintainer-side build artifact. It is not
part of the installed package and should not be confused with the user cache
created by `update_slcflights_data()`.

You can also rebuild a subset of years, for example:

```r
source("data-raw/build_data.R")
build_slc_data(years = 2019:2024)
```

Because rebuilding requires large downloads and substantial disk space, most
users should work with the installed files and, when needed, use
`update_slcflights_data()` for post-June-2024 updates.

## Data provenance

This package repackages a Salt Lake City-focused subset of the flight data used
in the 2025 ASA Data Expo Challenge. The challenge centered on analysis and
visualization of U.S. commercial flight arrival and departure records, while
the underlying data ultimately come from BTS TranStats On-Time Performance
data.

The packaged coordinate table is derived from the BTS TranStats Master
Coordinate support table and contains the airport sequence identifiers needed
to enrich the flight records with airport latitude, longitude, and date-bounded
airport metadata.
