---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# slcflights

<!-- badges: start -->
<!-- badges: end -->

`slcflights` provides Salt Lake City-focused subsets of the Bureau of
Transportation Statistics (BTS) TranStats **On-Time: Reporting Carrier
On-Time Performance** data.

The package is designed for analyses where Salt Lake City International
Airport is central to the question. It includes flights where Salt Lake City
appears as the scheduled origin or destination, plus a separate diversion-only
data grouping for flights where Salt Lake City appears only as a diversion
airport.

The installed package data cover **October 1, 1987 through June 30, 2024**.
Because the endpoint years are partial, the cleanest full-year span in the
installed data is **1988 through 2023**.

## Installation

You can install the development version of `slcflights` from GitHub:

``` r
# install.packages("pak")
pak::pak("AlKulAK/slcflights")
```

## Data sources and coverage

The flight records come from BTS TranStats **On-Time: Reporting Carrier
On-Time Performance** data.

The installed historical files are derived from the 1987--2024 Parquet files
released for the 2025 ASA Data Expo Challenge. Those files contain national
flight records. The `slcflights` build workflow filters them to Salt Lake
City-related records, removes globally empty columns, splits main records from
diversion-only records, and enriches airport sequence identifiers with
coordinate metadata.

For months after June 2024, `update_slcflights_data()` can download monthly
BTS files directly from TranStats and build a compatible local user cache. The
installed package files are never modified.

## What the package includes

`slcflights` includes two flight-data groupings:

- **main records**: flights where Salt Lake City's BTS airport ID appears in
  `OriginAirportID` or `DestAirportID`
- **diversion-only records**: flights where Salt Lake City's BTS airport ID
  appears in one of `Div1AirportID` through `Div5AirportID`, but not in
  `OriginAirportID` or `DestAirportID`

The package also includes:

- an airport coordinate table derived from the BTS TranStats Master Coordinate
  support table
- a field dictionary describing selected flight-record and coordinate-table
  columns
- cache-aware readers that combine installed data with compatible local cached
  data when a cache is active

## Quick start

Load the package:

```{r}
library(slcflights)
```

Check the available years:

```{r}
available_years("main")
available_years("div")
```

Open main records lazily with Arrow when you want to inspect the data before
materializing it:

```{r}
ds_main <- open_main(1988:1989)
ds_main
```

Read one year into memory when you want an ordinary data frame:

```{r}
x_1988 <- read_year_main(1988)
dim(x_1988)
head(x_1988[, c(
  "FlightDate",
  "Reporting_Airline",
  "OriginCityName",
  "DestCityName"
)])
```

Read several years at once when you want a single row-bound data frame:

```{r}
x_main <- read_main(1988:1989)
dim(x_main)
unique(x_main$Year)
```

Read the airport coordinate table directly:

```{r}
coords <- read_coords()
dim(coords)
names(coords)[1:8]
```

Read the field dictionary when you want column descriptions and field-presence
metadata:

```{r}
fields <- read_field_dictionary()
dim(fields)
names(fields)
```

## Main function families

### Discover available data

Use `available_years()` to list years available for main records or
diversion-only records.

```{r}
available_years("main")
available_years("div")
```

The endpoint years in the installed data are partial. Use 1988 through 2023
for analyses that require complete calendar years and do not use a local cache.

### Work lazily with Parquet files

Use `open_main()` and `open_div()` when you want an Arrow dataset instead of an
in-memory data frame.

```{r}
ds <- open_main(1988:1989)
ds
```

This is usually the best entry point for larger workflows where you want to
filter rows or select columns before collecting data into memory.

### Read data into memory

Use `read_year_main()` and `read_year_div()` for a single year.

```{r}
x <- read_year_main(1988)
head(x)
```

Use `read_main()` and `read_div()` for one, many, or all available years.

```{r}
x <- read_main(1988:1989)
dim(x)
```

Diversion-only records are packaged separately because they answer a different
question: cases where Salt Lake City matters because of a diversion event, not
because it was the scheduled origin or destination.

```{r, eval = FALSE}
div <- read_year_div(2015)
head(div)
```

### Read metadata tables

Use `read_coords()` when you want the standalone airport coordinate table.

```{r}
coords <- read_coords()
head(coords)
```

Use `read_field_dictionary()` when you want column descriptions, source-table
information, and field-presence metadata for main records, diversion-only
records, and the coordinate table.

```{r}
fields <- read_field_dictionary()
head(fields)
```

The flight Parquet files are already enriched with latitude, longitude, and
other date-bounded airport metadata. In most workflows, you do not need a
separate coordinate join.

## Updating local data

The installed package data end on **June 30, 2024**. To extend the available
data with newer BTS monthly releases, build a local cache:

```{r, eval = FALSE}
update_slcflights_data()
```

By default, `update_slcflights_data()` checks BTS for the latest available
monthly data and downloads every available month after June 2024. You can also
request an explicit endpoint:

```{r, eval = FALSE}
# One additional month of data
update_slcflights_data(until = "2024-07") 

# Six additional months of data
update_slcflights_data(until = c(2024, 12))
```

Local updates always begin with July 2024 and form a consecutive extension of
the installed data. The cache is stored in the user cache directory returned by
`tools::R_user_dir("slcflights", "cache")`.

Inspect the active cache with:

```{r, eval = FALSE}
slcflights_cache_info()
```

Remove the local cache with:

```{r, eval = FALSE}
clear_slcflights_cache()
```

After a compatible cache is active, the ordinary reader functions use it
automatically. For example, after updating through July 2024,
`read_year_main(2024)` reads the installed January-through-June records and the
cached July records as one combined data frame.

## Rebuilding the packaged data

Most users should **not** rebuild the packaged data. They should use the installed
files and, when needed, `update_slcflights_data()` for post-June-2024 monthly
updates.

The maintainer build workflow lives in `data-raw/build_data.R`. From a source
checkout, maintainers can rebuild the bundled Parquet files and coordinate CSV
with:

```{r, eval = FALSE}
source("data-raw/build_data.R")
build_slc_data()
```

By default, `build_slc_data()` rebuilds years 1987 through 2024, filters the
national source data to Salt Lake City-related records, splits main and
diversion-only files, reduces the BTS Master Coordinate table, enriches the
Parquet files with airport metadata, and writes final package data under
`inst/extdata/`.

The temporary build directory is a maintainer-side artifact. It is not part of
the installed package and is separate from the local user cache created by
`update_slcflights_data()`.

## Learn more

See `vignette("slcflights")` for a fuller walkthrough of the package workflow,
including coverage, lazy inspection, in-memory reads, airport metadata,
diversion-only records, and local cache updates.
