---
title: "Getting started with slcflights"
output: rmarkdown::html_vignette
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Getting started with slcflights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

## Overview

`slcflights` provides Salt Lake City-focused subsets of the Bureau of
Transportation Statistics (BTS) TranStats **On-Time: Reporting Carrier
On-Time Performance** data.

The installed package data cover **October 1, 1987 through June 30, 2024**.
The first and last calendar years are therefore partial years. The package keeps
those endpoint years because they are part of the historical source data used to
build the package, but analyses that require complete calendar years should
treat 1987 and 2024 with care.

This vignette uses the package for a Salt Lake City-centered adaptation of an
idea from Wickham's ASA 2009 Data Expo paper [@wickham2011asa]. Wickham used
the national on-time flight data to compare flight volumes on weekdays and
selected holidays after normalizing by each year's average day. Here we apply
the same normalization idea to `slcflights`, first for flights where Salt Lake
City was the scheduled origin or destination, and then for diversion-only
records where Salt Lake City appears only as a diversion airport.

The goal is not to reproduce Wickham's national figure. The goal is to use the
same analytical idea in a different setting: a Salt Lake City-centered package
with two distinct flight-data groupings.

## Loading the package

We begin by loading `slcflights` and all the other packages used in this vignette.

```{r setup}
library(slcflights)
library(dplyr)
library(grid)
library(ggplot2)
library(lubridate)
```

## Data provenance

The underlying flight records come from the BTS TranStats **On-Time:
Reporting Carrier On-Time Performance** table:

[https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FGJ&QO_fu146_anzr=b0-gvzr](https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FGJ&QO_fu146_anzr=b0-gvzr)

The BTS field documentation for this table is available at:

[https://www.transtats.bts.gov/Fields.asp?gnoyr_VQ=FGJ](https://www.transtats.bts.gov/Fields.asp?gnoyr_VQ=FGJ)

The installed historical files in `slcflights` are derived from the 1987--2024
Parquet files distributed for the 2025 ASA Data Expo Challenge:

[https://community.amstat.org/jointscsg-section/dataexpo/dataexpo202136710288](https://community.amstat.org/jointscsg-section/dataexpo/dataexpo202136710288)

Airport coordinate metadata come from the BTS TranStats Master Coordinate
support table in the Aviation Support Tables family:

[https://transtats.bts.gov/DL_SelectFields.aspx?QO_fu146_anzr=N8vn6v10+f722146+gnoyr5&gnoyr_VQ=FLL](https://transtats.bts.gov/DL_SelectFields.aspx?QO_fu146_anzr=N8vn6v10+f722146+gnoyr5&gnoyr_VQ=FLL)

The package build workflow transforms the national source files into Salt Lake
City-focused data by filtering to records involving Salt Lake City's BTS airport
ID, splitting scheduled origin/destination records from diversion-only records,
removing globally empty columns, sorting records, reducing the coordinate table
to airports referenced by retained records, and enriching the Parquet files
with coordinate metadata.

## What data are available?

The package has two flight-data groupings.

The **main** records are flights where Salt Lake City's BTS airport ID appears
in `OriginAirportID` or `DestAirportID`.

The **diversion-only** records are flights where Salt Lake City's BTS airport
ID appears in one of `Div1AirportID` through `Div5AirportID`, but not in
`OriginAirportID` or `DestAirportID`.

```{r available-years}
available_years("main")
available_years("div")
```

The main records are available for the full installed historical range,
including the partial endpoint years 1987 and 2024. Diversion-only records
begin later because the relevant diversion fields are not present throughout
the full historical period.

## Looking up fields

Before building an analysis, it is useful to inspect the field dictionary. The
dictionary describes selected fields and records whether each field appears in
main records, diversion-only records, or the coordinate table.

```{r field-dictionary}
fields <- read_field_dictionary()

fields |>
  filter(field %in% c("Year", "FlightDate", "Flights")) |>
  select(field, group, source, description, main_presence, div_presence)
```

For this vignette, the essential fields are `Year`, `FlightDate`, and
`Flights`. We use `Year` to normalize within each year, `FlightDate` to classify
days by weekday or holiday, and `Flights` to count flights.

## Calendar helpers

The following helpers create a lookup table of selected U.S. federal holidays. The
set of holidays follows the spirit of Wickham's weekday/holiday comparison:
ordinary weekdays are compared with selected holidays after normalizing by each
year's average day.

```{r calendar-helpers}
nth_weekday_of_month <- function(year, month, weekday, n) {
  # Return the date of the nth requested weekday in a given month.
  # weekday uses lubridate's week_start = 1 convention:
  # Monday = 1, Tuesday = 2, ..., Sunday = 7.
  first_day <- as.Date(sprintf("%04d-%02d-01", year, month))
  first_wday <- lubridate::wday(first_day, week_start = 1)

  first_day + ((weekday - first_wday) %% 7) + 7L * (n - 1L)
}

last_weekday_of_month <- function(year, month, weekday) {
  # Return the date of the last requested weekday in a given month.
  # weekday uses lubridate's week_start = 1 convention:
  # Monday = 1, Tuesday = 2, ..., Sunday = 7.
  if (month == 12L) {
    next_month <- as.Date(sprintf("%04d-01-01", year + 1L))
  } else {
    next_month <- as.Date(sprintf("%04d-%02d-01", year, month + 1L))
  }

  last_day <- next_month - 1L
  last_wday <- lubridate::wday(last_day, week_start = 1)

  last_day - ((last_wday - weekday) %% 7)
}

make_holiday_table <- function(years) {
  # Build a Year-Date-category lookup table for the selected holidays included
  # in the weekday/holiday comparison.
  years <- sort(unique(as.integer(years)))

  holiday_names <- c(
    "New Year's Day",
    "Martin Luther King Jr. Day",
    "Presidents' Day",
    "Memorial Day",
    "Independence Day",
    "Labor Day",
    "Veterans Day",
    "Thanksgiving Day",
    "Christmas Day"
  )

  holiday_dates <- c(
    as.Date(sprintf("%04d-01-01", years)),
    nth_weekday_of_month(years, 1L, 1L, 3L),
    nth_weekday_of_month(years, 2L, 1L, 3L),
    last_weekday_of_month(years, 5L, 1L),
    as.Date(sprintf("%04d-07-04", years)),
    nth_weekday_of_month(years, 9L, 1L, 1L),
    as.Date(sprintf("%04d-11-11", years)),
    nth_weekday_of_month(years, 11L, 4L, 4L),
    as.Date(sprintf("%04d-12-25", years))
  )

  data.frame(
    Year = rep(years, times = length(holiday_names)),
    FlightDate = holiday_dates,
    category = factor(
      rep(holiday_names, each = length(years)),
      levels = c(
        "Monday",
        "Tuesday",
        "Wednesday",
        "Thursday",
        "Friday",
        "Saturday",
        "Sunday",
        "Martin Luther King Jr. Day",
        "Presidents' Day",
        "Memorial Day",
        "Independence Day",
        "Labor Day",
        "Veterans Day",
        "Thanksgiving Day",
        "Christmas Day",
        "New Year's Day"
      )
    )
  )
}
```

## A Salt Lake City adaptation of Wickham's comparison

Wickham's normalization idea is useful because it removes year-to-year traffic
growth from the comparison. Instead of asking how many flights occurred on a
particular holiday or weekday, we ask how each category compares with that
year's average day.

A value of 100 means that the category is equal to that year's average day.
Values below 100 indicate fewer flights than an average day in that year, and
values above 100 indicate more flights than an average day in that year.

The same calculation is applied to both `slcflights` data groupings. That makes
the contrast meaningful: main records describe scheduled SLC service, while
diversion-only records describe rare operational disruptions.

```{r analysis-functions}
make_day_type_data <- function(type = c("main", "div"), years = NULL) {
  type <- match.arg(type)

  # Use all available years for the requested data grouping unless the caller
  # supplies a smaller set of years.
  if (is.null(years)) {
    years <- sort(unique(slcflights::available_years(type)))
  } else {
    years <- sort(unique(as.integer(years)))
  }

  # Choose the lazy reader for the requested data grouping. Main records are
  # scheduled SLC origin/destination flights; diversion-only records are flights
  # where SLC appears only as a diversion airport.
  open_fun <- switch(type,
    main = slcflights::open_main,
    div = slcflights::open_div
  )

  # Open the Parquet files lazily, select only the fields needed for this
  # analysis, and collapse the records to one total flight count per calendar
  # day. The collect() call happens after aggregation, so the vignette brings
  # only the smaller daily summary into memory.
  daily <- open_fun(years) |>
    dplyr::select(Year, FlightDate, Flights) |>
    dplyr::group_by(Year, FlightDate) |>
    dplyr::summarise(daily_flights = sum(Flights), .groups = "drop") |>
    dplyr::collect() |>
    dplyr::mutate(
      FlightDate = as.Date(FlightDate),
      weekday = factor(
        lubridate::wday(FlightDate, week_start = 1),
        levels = 1:7,
        labels = c(
          "Monday",
          "Tuesday",
          "Wednesday",
          "Thursday",
          "Friday",
          "Saturday",
          "Sunday"
        )
      )
    )

  # Compute each year's average daily flight count. This value is the
  # denominator for the Wickham-style normalization.
  year_avg <- daily |>
    dplyr::group_by(Year) |>
    dplyr::summarise(avg_daily_year = mean(daily_flights), .groups = "drop")

  # For ordinary weekdays, compute one average daily flight count per weekday
  # within each year.
  weekday_stats <- daily |>
    dplyr::group_by(Year, category = weekday) |>
    dplyr::summarise(
      avg_daily_category = mean(daily_flights),
      .groups = "drop"
    )

  # For holidays, each year contributes the daily count on that holiday. Holiday
  # dates outside the installed October 1987-June 2024 data window drop out
  # naturally in the inner join.
  holiday_stats <- daily |>
    dplyr::inner_join(
      make_holiday_table(years),
      by = c("Year", "FlightDate")
    ) |>
    dplyr::group_by(Year, category) |>
    dplyr::summarise(
      avg_daily_category = mean(daily_flights),
      .groups = "drop"
    )

  # Combine weekday and holiday summaries. Each output row is one
  # Year-category summary, expressed as a percentage of that year's average
  # daily flight count.
  dplyr::bind_rows(weekday_stats, holiday_stats) |>
    dplyr::left_join(year_avg, by = "Year") |>
    dplyr::mutate(
      pct_of_avg_day = 100 * avg_daily_category / avg_daily_year,
      category = factor(
        as.character(category),
        levels = c(
          "Monday",
          "Tuesday",
          "Wednesday",
          "Thursday",
          "Friday",
          "Saturday",
          "Sunday",
          "Martin Luther King Jr. Day",
          "Presidents' Day",
          "Memorial Day",
          "Independence Day",
          "Labor Day",
          "Veterans Day",
          "Thanksgiving Day",
          "Christmas Day",
          "New Year's Day"
        )
      )
    )
}

plot_day_type <- function(
  type = c("main", "div"),
  ordering = c("chronological", "median_desc")
) {
  type <- match.arg(type)
  ordering <- match.arg(ordering)

  plot_df <- make_day_type_data(type = type)

  # Use chronological ordering to preserve the weekday/holiday order, or order
  # by median percentage to make high- and low-volume categories easier to see.
  if (ordering == "chronological") {
    plot_df <- plot_df |>
      dplyr::mutate(
        category = factor(
          as.character(category),
          levels = rev(levels(category))
        )
      )
  } else if (ordering == "median_desc") {
    category_order <- plot_df |>
      dplyr::group_by(category) |>
      dplyr::summarise(med = median(pct_of_avg_day), .groups = "drop") |>
      dplyr::arrange(dplyr::desc(med)) |>
      dplyr::pull(category) |>
      as.character()

    plot_df <- plot_df |>
      dplyr::mutate(
        category = factor(
          as.character(category),
          levels = rev(category_order)
        )
      )
  }

  # Add sample sizes to the y-axis labels. n is the number of annual category
  # summaries in each boxplot. Holiday sample sizes can be smaller because 1987
  # and 2024 are partial years; diversion-only records also begin later.
  sample_size_df <- plot_df |>
    dplyr::count(category, name = "n") |>
    dplyr::mutate(
      label = paste0(as.character(category), "\nn = ", n)
    )

  category_labels <- sample_size_df$label
  names(category_labels) <- as.character(sample_size_df$category)

  caption_text <- paste(
    "n is the number of annual category summaries in each boxplot;",
    "1987 and 2024 are partial years."
  )

  # Main records have a tighter range; diversion-only records are much more
  # variable, so they need a wider x-axis.
  x_breaks <- switch(type,
    main = seq(40, 140, by = 10),
    div = seq(20, 200, by = 10)
  )

  x_limits <- switch(type,
    main = c(40, 140),
    div = c(20, 200)
  )

  ggplot2::ggplot(
    plot_df,
    ggplot2::aes(x = pct_of_avg_day, y = category)
  ) +
    ggplot2::geom_vline(xintercept = 100, linetype = 2) +
    ggplot2::geom_boxplot() +
    ggplot2::scale_y_discrete(labels = category_labels) +
    ggplot2::scale_x_continuous(breaks = x_breaks) +
    ggplot2::coord_cartesian(xlim = x_limits) +
    ggplot2::labs(
      x = "Percentage of flights relative to average day",
      y = NULL,
      caption = caption_text
    )
}

draw_plot_pair <- function(ordering = c("chronological", "median_desc")) {
  ordering <- match.arg(ordering)

  # Build two ggplot objects and draw them on one graphics page. This avoids
  # relying on HTML layout behavior for two separate images.
  p_main <- plot_day_type("main", ordering = ordering)
  p_div <- plot_day_type("div", ordering = ordering)

  grid::grid.newpage()
  grid::pushViewport(
    grid::viewport(
      layout = grid::grid.layout(nrow = 1, ncol = 2)
    )
  )

  print(
    p_main,
    vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1)
  )

  print(
    p_div,
    vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2)
  )
}
```

The following pair of plots shows the same weekday and holiday categories in calendar
order. The left plot uses main records, where Salt Lake City was the scheduled
origin or destination. The right plot uses diversion-only records, where Salt
Lake City appears only because of a diversion. Each boxplot summarizes annual
category values, expressed as a percentage of that year's average daily flight
count.

The main and diversion-only plots use different x-axis ranges because the
diversion-only records are much more variable.

```{r paired-chronological, fig.width = 12, fig.height = 6, out.width = "100%", fig.align = "center"}
draw_plot_pair(ordering = "chronological")
```

The next pair shows the same summaries ordered by median percentage of an
average day. This ordering makes the contrast easier to read: the main-record
plot reflects a regular scheduled-service rhythm, while the diversion-only plot
is much more variable.

```{r paired-median-desc, fig.width = 12, fig.height = 6, out.width = "100%", fig.align = "center"}
draw_plot_pair(ordering = "median_desc")
```

For main SLC records, the weekday categories cluster close to the average-day
baseline. Saturday is lower than the ordinary weekday categories, and several
major holidays tend to sit below the baseline. This is the expected behavior
for scheduled passenger service: ordinary weekdays reflect regular airline
schedules, while some holidays alter demand and schedule patterns.

The diversion-only plots are much less stable. This is not a defect in the
package or in the method. It reflects the nature of the data grouping.
Diversion-only records are rare events, and some holidays appear in only a few
annual summaries. The same normalization remains useful, but it should be read
as a comparison of operational volatility rather than a regular travel-demand
pattern.

## Local updates

The installed package data end on **June 30, 2024**. To extend the available
data with newer BTS monthly releases, run:

```{r update-data, eval = FALSE}
update_slcflights_data()
```

By default, `update_slcflights_data()` checks BTS for the latest available
monthly data and downloads every available month after June 2024. You can also
request an explicit endpoint:

```{r update-data-explicit, eval = FALSE}
# One additional month of data
update_slcflights_data(until = "2024-07")

# Six additional months of data
update_slcflights_data(until = c(2024, 12))
```

Local updates always begin with July 2024 and form a consecutive extension of
the installed data. The update process stores data in the user cache directory
returned by `tools::R_user_dir("slcflights", "cache")`. The installed package
files are not modified.

You can inspect the active cache with `slcflights_cache_info()`. This is useful
after an update because it reports whether a compatible local cache is active
and which years are available from it.

If you want to discard the local cache and return to using only the data
installed with the package, run:

```{r clear-cache, eval = FALSE}
clear_slcflights_cache()
```

After a compatible cache is active, ordinary reader functions use it
automatically. For example, after updating through July 2024,
`read_year_main(2024)` reads the installed January-through-June records and the
cached July records as one combined data frame.

## Rebuilding the bundled data

Most users should **not** rebuild the bundled data. They should use the installed
files and, when needed, `update_slcflights_data()` for post-June-2024 monthly
updates.

The maintainer build workflow lives in `data-raw/build_data.R`. From a source
checkout, maintainers can rebuild all bundled data artifacts with:

```{r rebuild-data, eval = FALSE}
source("data-raw/build_data.R")
build_slc_data()
```

By default, `build_slc_data()` rebuilds all bundled data artifacts. It rebuilds
years 1987 through 2024, filters the national source data to Salt Lake
City-related records, splits main and diversion-only files, reduces the BTS
Master Coordinate table, enriches the Parquet files with airport metadata,
rebuilds the field dictionary, and writes final package data under
`inst/extdata/`.

No source data files need to be present before running this workflow. The
historical Parquet source files are downloaded into a temporary build directory,
and the BTS Master Coordinate table is downloaded automatically if
`data-raw/T_MASTER_CORD.csv` is absent.

The temporary build directory is a maintainer-side artifact. It is not part of
the installed package and is separate from the local user cache created by
`update_slcflights_data()`.

## Summary

This vignette used `slcflights` to adapt Wickham's weekday/holiday
normalization idea to Salt Lake City-centered flight records. The main records
show regular scheduled-service patterns, while the diversion-only records are
sparser and more volatile. The contrast illustrates why `slcflights` keeps
these two data groupings separate.

## References {#refs}
