---
title: "Getting started with slcflights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with slcflights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

## Overview

`slcflights` builds a local Salt Lake City-focused database from the Bureau of
Transportation Statistics (BTS) TranStats **On-Time: Reporting Carrier
On-Time Performance** data.

The package does not install flight records. Instead, users build a local
database with `build_slcflights_db()`. The local database begins with October
1987 and extends through the endpoint chosen by the user. The endpoint is the
last year or month included in the local database.

Because the database begins with October 1987, the first calendar year is
partial. The endpoint year may also be partial.

`slcflights` keeps two flight-record groupings:

- **main records**, where Salt Lake City's BTS airport ID appears as the
  scheduled origin or destination; and
- **diversion-only records**, where Salt Lake City's BTS airport ID appears only
  as a diversion airport.

This vignette shows the basic workflow: build a local database, check its
status, inspect available years, read records, use metadata tables, and manage
the local database.

## Before building the database

Building or updating the local database can take substantial time and disk
space. This is true even for relatively small date ranges because `slcflights`
downloads and processes monthly national BTS source files before writing the
local Salt Lake City-focused database.

When a monthly BTS PREZIP file is available, `slcflights` downloads that static
ZIP file directly. When a PREZIP file is not available, the package
automatically uses the TranStats all-fields export for the same BTS table. That
fallback can be slower and can require more temporary storage because the
package has to request and process a generated export.

For a first build, start with a small endpoint if you want to check your setup
before committing to a larger database.

```{r small-build-example, eval = FALSE}
build_slcflights_db(until = "1989-12", confirm = TRUE)
```

The database is stored in the user cache directory returned by
`tools::R_user_dir("slcflights", "cache")`. Building or updating the database
does not modify installed package files.

## Loading packages

This vignette uses `slcflights` for data access and `dplyr` for small examples.

```{r setup}
library(slcflights)
library(dplyr)
```

## Data provenance

The underlying flight records come from the BTS TranStats **On-Time:
Reporting Carrier On-Time Performance** table:

[https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FGJ&QO_fu146_anzr=b0-gvzr](https://www.transtats.bts.gov/DL_SelectFields.aspx?gnoyr_VQ=FGJ&QO_fu146_anzr=b0-gvzr)

The BTS field documentation for this table is available at:

[https://www.transtats.bts.gov/Fields.asp?gnoyr_VQ=FGJ](https://www.transtats.bts.gov/Fields.asp?gnoyr_VQ=FGJ)

Airport coordinate metadata come from the BTS TranStats Master Coordinate
support table in the Aviation Support Tables family:

[https://transtats.bts.gov/DL_SelectFields.aspx?QO_fu146_anzr=N8vn6v10+f722146+gnoyr5&gnoyr_VQ=FLL](https://transtats.bts.gov/DL_SelectFields.aspx?QO_fu146_anzr=N8vn6v10+f722146+gnoyr5&gnoyr_VQ=FLL)

Airline ID metadata come from the BTS TranStats DOT_ID_Reporting_Airline lookup
table:

[https://www.transtats.bts.gov/Download_Lookup.asp?Y11x72=Y_NVeYVaR_VQ](https://www.transtats.bts.gov/Download_Lookup.asp?Y11x72=Y_NVeYVaR_VQ)

The local database build workflow downloads monthly BTS source files, filters
the national records to flights involving Salt Lake City's BTS airport ID,
splits scheduled origin/destination records from diversion-only records, sorts
records, reduces the coordinate and Airline ID lookup tables to metadata rows
referenced by retained records, and writes Parquet files already enriched with
coordinate metadata and airline names.

## Building and checking a local database

Use `build_slcflights_db()` to create a local database. The endpoint can be a
year, a year-month string, `c(year, month)`, or `"latest"`. The `"latest"`
endpoint means the latest month the package can resolve from BTS at build or
update time.

```{r build-database, eval = FALSE}
build_slcflights_db(until = "2024-12", confirm = TRUE)
```

Use `status_slcflights_db()` to check whether a local database is active and
which endpoint it contains.

```{r database-status, eval = FALSE}
status_slcflights_db()
```

After a database exists, reader functions use it automatically.

## Inspecting available data

The package has two flight-data groupings.

The **main** records are flights where Salt Lake City's BTS airport ID appears
in `OriginAirportID` or `DestAirportID`.

The **diversion-only** records are flights where Salt Lake City's BTS airport
ID appears in one of `Div1AirportID` through `Div5AirportID`, but not in
`OriginAirportID` or `DestAirportID`.

Use `available_years()` to inspect the active local database. By default,
`available_years()` returns years for main records. Use `type = "div"` to
inspect diversion-only years.

```{r available-years, eval = FALSE}
available_years()
available_years(type = "div")
```

Main-record years reflect the years covered by the active local database.
Diversion-only records begin in October 2008, so their available years will always be a
smaller set.

## Reading records

Use `open_main()` and `open_div()` when you want to work lazily with Arrow
datasets. This is useful when you want to filter, select, or aggregate before
collecting data into memory.

```{r lazy-readers, eval = FALSE}
main_ds <- open_main(1988:1989)
div_ds <- open_div(2009:2010)
```

Use `read_year_main()` and `read_year_div()` to read one year into memory.

```{r year-readers, eval = FALSE}
main_1988 <- read_year_main(1988)
div_2009 <- read_year_div(2009)
```

Use `read_main()` and `read_div()` to read one year, several years, or all
available years.

```{r multi-year-readers, eval = FALSE}
main_early <- read_main(1988:1989)
div_recent <- read_div(2018:2019)
```

The examples above assume that the active local database covers the requested
years. Diversion-only examples require a database endpoint of October 2008 or
later.

Calling `read_main()` or `read_div()` without a year argument reads all
available years for that grouping into memory. For larger workflows, prefer the
lazy readers and collect only the summarized result.

Once records are in memory, ordinary data-frame tools can be used. For example,
the following code counts the most common origin-destination pairs in one year
of main records.

```{r origin-dest-example, eval = FALSE}
main_1988 |>
  count(Origin, Dest, sort = TRUE) |>
  slice_head(n = 10)
```

For larger queries, prefer the lazy readers and collect only the summarized
result.

```{r lazy-summary-example, eval = FALSE}
open_main(1988:1989) |>
  select(Year, Origin, Dest, Flights) |>
  group_by(Year, Origin, Dest) |>
  summarise(flights = sum(Flights), .groups = "drop") |>
  arrange(desc(flights)) |>
  collect()
```

## Reading metadata tables

The local database includes reduced airport coordinate and airline lookup
tables. These tables contain metadata rows referenced by retained flight
records.

Most analyses do not require manual joins to these tables. During the database
build, `slcflights` writes main and diversion Parquet files already enriched
with coordinate metadata and airline names. The standalone metadata readers are
mainly useful when you want to inspect the reduced lookup tables directly,
check identifiers, or support specialized analysis.

Use `read_coords()` to read airport coordinate metadata.

```{r coordinate-lookup, eval = FALSE}
coords <- read_coords()

coords |>
  select(AirportID, Display_Airport_Name, Airport_Lat, Airport_Lon) |>
  slice_head(n = 10)
```

Use `read_airlines()` to read airline lookup metadata.

```{r airline-lookup, eval = FALSE}
airlines <- read_airlines()

airlines |>
  select(
    DOT_ID_Reporting_Airline,
    Reporting_Airline_Name,
    Reporting_Airline_Lookup_Code
  ) |>
  slice_head(n = 10)
```

## Looking up fields

The field dictionary describes fields retained by `slcflights` and records
whether each field appears in main records, diversion-only records, or the
coordinate table.

```{r field-dictionary}
fields <- read_field_dictionary()

fields |>
  filter(
    field %in% c(
      "Year",
      "FlightDate",
      "Flights",
      "OriginAirportID",
      "DestAirportID",
      "DOT_ID_Reporting_Airline",
      "Reporting_Airline_Name"
    )
  ) |>
  select(field, group, source, description, main_presence, div_presence)
```

Unlike the flight-record readers, the field dictionary is available without
building a local database. It is a quick way to check field names before writing
a query.

## Managing the local database

Use `update_slcflights_db()` to extend an active database forward.

```{r extend-data, eval = FALSE}
update_slcflights_db(until = "latest", confirm = TRUE)
```

Updates are contiguous. If the active database ends with one month, the next
extension starts with the following month.

Use `delete_slcflights_db()` when you want to remove the active local database.

```{r delete-data, eval = FALSE}
delete_slcflights_db()
```

To rebuild from scratch, delete the active database and build a new one.

```{r rebuild-data, eval = FALSE}
delete_slcflights_db()
build_slcflights_db(until = "latest", confirm = TRUE)
```

The `data-raw/` scripts are development artifacts. The ordinary workflow is to
build, update, inspect, read, or delete the local database with the exported
package functions.

## Summary

This vignette showed the basic `slcflights` workflow: build a local database,
check its status, inspect available years, read records, use metadata tables,
look up field definitions, and manage the local database. Main records describe
scheduled SLC origin/destination service. Diversion-only records describe rarer
operational disruptions. Keeping these groupings separate makes scheduled SLC
service and diversion patterns easier to analyze separately.
