---
title: "slcflights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{slcflights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(slcflights)
library(arrow)
```

## Overview

`slcflights` packages Salt Lake City-focused U.S. flight records as Parquet
files, along with a standalone airport lookup table derived from BTS
coordinate data.

This vignette introduces the main workflow:

1. understand the package's time coverage
2. inspect which years are packaged
3. inspect packaged Parquet files lazily
4. read selected data into memory
5. access the standalone airport lookup table when you want airport metadata
   on its own

The package includes two flight-data groupings:

- **main records**: flights where Salt Lake City appears in the primary
  origin or destination airport fields
- **diversion-only records**: flights where Salt Lake City appears only in
  diversion airport fields and not in the primary origin or destination
  fields

## Coverage and an important caveat

The packaged flight data cover the period from **October 1, 1987** through
**June 30, 2024**, meaning that:

- **1987 is a partial year**
- **2024 is a partial year**

Therefore, while `available_years()` reports 1987 through 2024 as packaged
years, those endpoint years do **not** contain a full January-through-December
span of records.

This matters for year-over-year comparisons. If you want a run of full years,
the cleanest span is **1988 through 2023**.

You can see the partial-year behavior directly for 1987:

```{r}
x_1987 <- slcflights::read_year_main(1987)
range(x_1987$FlightDate)
```

## See what years are packaged

Start by checking which years are available.

```{r}
slcflights::available_years("main")
```

Diversion-only coverage does differ from main-record coverage, so inspect it
separately:

```{r}
slcflights::available_years("div")
```

Again, keep in mind that availability is not the same as full-year completeness.

## Inspect packaged Parquet files lazily

A good first step is often to open the packaged Parquet files lazily with
Arrow.

Use `open_main()` or `open_div()` when you want to inspect the packaged data
at a high level without immediately reading all selected rows into memory.

```{r}
ds_main <- slcflights::open_main(1988:1989)
ds_main
```

You can inspect the dataset schema directly:

```{r}
arrow::schema(ds_main)
```

This is a convenient way to understand the overall structure of the packaged
data before deciding what subset you want to materialize.

## Read data into memory when you are ready to work with it directly

Once you know what years you want and are ready to manipulate the data as an
ordinary in-memory object, use the `read_*()` helpers.

Use `read_year_main()` or `read_year_div()` for a single year:

```{r}
x_1988 <- slcflights::read_year_main(1988)

dim(x_1988)
names(x_1988)
```

Here is a very small subset of potentially useful columns:

```{r}
head(x_1988[, c(
  "FlightDate",
  "Reporting_Airline",
  "OriginCityName",
  "DestCityName"
)])
```

Use `read_main()` or `read_div()` when you want multiple years combined
row-wise into a single in-memory data frame.

```{r}
x_main <- slcflights::read_main(1988:1989)

dim(x_main)
unique(x_main$Year)
```

Likewise, use `read_div()` for diversion-only records.

```{r}
x_div <- slcflights::read_div(2015:2016)
dim(x_div)
```

In short:

- use `open_*()` to inspect and explore the packaged Parquet files lazily
- use `read_*()` to load the data you actually want to analyze in memory

## Airport coordinates are already in the flight records

In many workflows, you do not need a separate join just to recover airport
coordinates. The packaged flight Parquet files already include airport
location and date-bounded airport metadata associated with the flight record.

You can inspect the coordinate-related fields directly:

```{r}
grep("lat|lon", names(x_1988), value = TRUE, ignore.case = TRUE)
```

And you can inspect those columns in the in-memory data you read:

```{r}
coord_cols <- grep("Latitude|Longitude", names(x_1988), value = TRUE)
coord_cols
head(x_1988[, coord_cols, drop = FALSE])
```

This is often the simplest way to work with airport locations in analyses of
routes, origins, destinations, and diversions.

## Work with diversion-only records

Diversion-only records are packaged separately because they answer a different
question from the main origin/destination files.

Use them when Salt Lake City is relevant because of a diversion event rather
than because it was the scheduled origin or destination.

```{r}
x_div_2015 <- slcflights::read_year_div(2015)

dim(x_div_2015)
names(x_div_2015)
head(x_div_2015[, c(
  "FlightDate",
  "Origin",
  "Dest"
)])
```

## Access the airport lookup table directly

Although the flight Parquet files already include airport coordinates and
related airport metadata, `slcflights` also provides a standalone airport
lookup table.

This is useful when you want airport metadata on its own, without first
reading any flight records.

```{r}
coords <- slcflights::read_coords()

dim(coords)
names(coords)
head(coords)
```

Use `read_coords()` when you want a compact airport reference table keyed by
airport sequence identifier, or when you want to inspect the packaged lookup
data directly.

## Choosing among the main entry points

Use `available_years()` when you want to discover coverage.

Use `open_main()` or `open_div()` when you want to inspect the packaged data
lazily without immediately materializing it.

Use `read_year_main()` or `read_year_div()` when you want one year's data in
memory.

Use `read_main()` or `read_div()` when you want multiple years combined into a
single in-memory data frame.

Use `read_coords()` when you want the standalone airport lookup table as its
own data object.

## Rebuilding the packaged data

The package also preserves the data-build workflow in `data-raw/`.

```{r, eval = FALSE}
# NOT EXECUTED
source("data-raw/build_data.R")
build_slc_data()
```

By default, `build_slc_data()` rebuilds data for years 1987 through 2024,
uses Salt Lake City's BTS airport ID, caches downloaded source files during
the build, writes packaged outputs into `inst/extdata/`, and removes the
build cache when finished.

You can also rebuild a subset of years.

```{r, eval = FALSE}
# NOT EXECUTED
source("data-raw/build_data.R")
build_slc_data(years = 2019:2024)
```

Rebuilding requires large downloads and substantial disk space, so most users
should work directly with the packaged files instead.

## Summary

`slcflights` is built around packaged Parquet files for Salt Lake City-related
flight records and a standalone airport lookup table.

The most important usage caveat is that the package covers **October 1, 1987**
through **June 30, 2024**, so **1987 and 2024 are partial years**. For
analyses that require full calendar years, focus on **1988 through 2023**.

From there, the package gives you three main access patterns:

- `available_years()` to inspect coverage
- `open_*()` helpers for lazy inspection
- `read_*()` helpers for eager in-memory access

And when you want airport metadata on its own, `read_coords()` provides the
packaged lookup table directly.
